---
title: "OpenGWAS"
author: "Brian M. Schilder"
date: "`r Sys.Date()`"
csl: nature.csl
output:
  BiocStyle::html_document:
vignette: >
    %\VignetteIndexEntry{OpenGWAS} 
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
---

# Install packages

To install the dev branch of `MAGMA.Celltyping`, use:

```{r, eval=FALSE, include=TRUE}
if(!require("remotes")) install.packages("remotes")
if(!require("here")) install.packages("here")
if(!require("BiocManager")) install.packages("BiocManager")


if(!require("MungeSumstats")) {
  BiocManager::install("MungeSumstats", dependencies = TRUE)
}
if(!require("EWCE") || packageVersion("EWCE")<1.3.1){
  remotes::install_github("NathanSkene/EWCE@bschilder_dev")
} 
if(!require("MAGMA.Celltyping")) {
  remotes::install_github("neurogenomics/MAGMA_Celltyping@bschilder_dev")
}
```


```{r setup, include=TRUE}
library(MAGMA.Celltyping)
library(MungeSumstats)

library(ggplot2)
library(dplyr)

source(here::here("code/utils.R"))

save_dir <- here::here("data/GWAS_sumstats")
dir.create(save_dir, showWarnings = FALSE, recursive = TRUE)
```

# Munge sum stats

## Find sum stats

### Neurodegeneration traits

First, we will search for GWAS of several neurodegenerative disease traits.

```{r} 
traits1 <- c("Alzheimer","Parkinson","amyotrophic","frontotemporal dementia","dementia")
metagwas1 <- MungeSumstats::find_sumstats(traits = traits1)
```

### BMI traits

Next, we perform a second search for Body Mass Index (BMI)-related traits.
We are querying these separately because we want to apply some additional filtering such that we only include GWAS with the greatest sample sizes from the UK Biobank, as well as exclude GWAS of "weight *loss*" or "weight *change*". 

```{r}
traits2 <- c("body mass index","standing height","weight")
metagwas2 <- MungeSumstats::find_sumstats(traits = traits)
metagwas2 <- filter_traits(meta = metagwas2, 
                            traits = traits, 
                            consortia = c("Neale Lab","UK Biobank"),
                            exclusion_terms = c("loss","change"),
                            startswith_only = TRUE,
                            topn = 1) 
```

### Sampling of neuro/immuno/cardio traits

```{r}
subcategories3 <- c("neurological","Immune","cardio")
metagwas3 <- MungeSumstats::find_sumstats(subcategories = subcategories3)
metagwas3 <- filter_traits(meta = metagwas3,  
                           group_var = "subcategory",
                           topn = 100) 
```

## All UKB traits

```{r}
meta4  <- MungeSumstats::find_sumstats(consortia = c("Neale Lab","UK Biobank"))
```

Finally, we can merge the metadata from our two GWAS searches.
We use th `assign_trait_groups` functions to automatically assign mutually-exclusive groups to each of the traits based on our search terms. 
**NOTE**: You may want to double check the "trait_group" assignments, as these may not always be what you had in mind; e.g. "frontotemporal dementia (FTD)" might get assigned to the generic "dementia" group instead of the more appropriate "frontotemporal dementia" group simply due to the order in which groups were assigned.

```{r}
metagwas <- rbind(metagwas1,metagwas2,metagwas3, fill=TRUE)
traits <- c(traits1,traits2)
meta <- assign_trait_groups(meta = metagwas, 
                            traits = traits)
knitr::kable(meta)
```

## Import sum stats

`MungeSumstats::import_sumstats` already supports parallelisation 
across multiple GWAS (by setting `nThread` >1). 
However, the log files would all 
overwrite one another. 

So we instead parellelise this step ourselves,
saving each summary stats file in it's own folder with a 
subdirectory for its respective log files.

```{r}
munged_files <- list_sumstats(save_dir = save_dir)
#### First, let's remove incomplete files.  ####
# bad_files <- munged_files[file.info(munged_files)$size <100]
# file.remove(bad_files)
gwas_paths <- parallel::mclapply(seq_len(nrow(meta)), function(i){
  MAGMA.Celltyping:::message_parallel(i," ======== >")
  id <- meta$id[i]
  if(id %in% names(munged_files)){
    MAGMA.Celltyping:::message_parallel(
      "Skipping: Pre-existing file for ",id)
    return(munged_files[id])
  } else {
     id_dir <- file.path(save_dir,id)
    dir.create(id_dir, showWarnings = FALSE, recursive = TRUE)
    log_dir <- file.path(id_dir,"logs")
    dir.create(log_dir, showWarnings = FALSE, recursive = TRUE)
    gwas <- tryCatch(expr = {
      MungeSumstats::import_sumstats(ids = id, 
                                     save_dir = id_dir, 
                                     nThread = 1, 
                                     parallel_across_ids = FALSE,
                                     force_new_vcf = FALSE,
                                     force_new = FALSE,
                                     #### Record logs
                                     log_folder_ind = TRUE,
                                     log_mungesumstats_msgs = TRUE,
                                     log_folder = log_dir)
    }, error = function(x){NA})
    return(setNames(gwas, id))
  }
}) 
 
```


# Create metadata file 

```{r}
meta <- gather_metadata(save_dir = save_dir)
```

# Map SNPs to genes 

```{r}
gene_files <- list_snps_to_genes_files(save_dir = save_dir)

magma_files <- parallel::mclapply(seq(1,nrow(meta)), function(i){
  MAGMA.Celltyping:::message_parallel("----- ",i," : ",
                                      meta$id[i]," -----")
  if(meta$id[i] %in% names(gene_files)) {
     MAGMA.Celltyping:::message_parallel("Skipping: Precomputed .genes.out file detected.")
    return(gene_files[[meta$id[i]]])
  } else{
     tryCatch(expr = {
       MAGMA.Celltyping::map_snps_to_genes(path_formatted = meta$munged_path[i],
                                      genome_build = meta$build_inferred[i],
                                      N = meta$N[i],
                                      upstream_kb = 35,  
                                      downstream_kb = 10, 
                                      force_new = FALSE)
     }, error = function(e) NULL)
  }
}) %>% `names<-`(meta$id)
```

# Gather snps2genes files

Now that all the GWAS have been mapped from SNPs to genes, 
we can copy the smaller mapping files into a folder that will 
get pushed to GitHub (i.e. isn't included in the *.gitignore* file).

```{r}

gene_files2 <- copy_snps_to_genes_files(search_dir = save_dir,
                                        save_dir = here::here("data/MAGMA_Files"), 
                                        overwrite = FALSE)

#### Find smallest files for demo purposes ####
# f <- unlist(gene_files2)
# data.frame(file=f,
#            size=file.info(f)$size) %>% dplyr::arrange(size)
```

## Re-save metadata

Add the paths of the gene mapping files to the metadata and resave it.

We will also add a URL that can be used to download the files. 

```{r}
meta <- gather_metadata(save_dir = save_dir)
```


# Construct gene x trait matrix

Can construct gene x trait matrix from gene-level MAGMA results 
across multiple GWAS traits.

```R
magma_matrix <- phenomix::gene_trait_matrix(magma_files = magma_files,
                                            metric = "ADJ_ZSTAT", 
                                            save_path = file.path(save_dir,"OpenGWAS_neurodegenGWAS_ADJZSTAT.rds"),
                                            mc.cores = 10) 
```


# Session Info

<details>

```{r}
utils::sessionInfo()
```

</details>
